---
title: 'PPAS Challenge: Fit a GLM'
output: 
  html_document:
    toc: True
---

## Background
In this challenge, we use data from the Titanic wreck in an attempt to "predict" the probability of surviving based on key traits. Logistic generalized linear models (GLMs) are used to predict or estimate probabilities.

## Goals
To practice using R's code to fit and review logistic GLMs, and to get a feel for what the coefficients mean.

## Load data and packages
```{r Load, warning = F, message = F}
library(car)

# from the "car" package datasets
titanic <- TitanicSurvival 

# or from the repository
titanic <- read.csv("TitanicSurvival.csv",
                    row.names = 1)
```

## Review data
```{r DataReview}
head(titanic)
summary(titanic)
```

## Challenges
### 1) Fit GLM
Using the glm() function, build a logistic GLM to predict/estimate passenger survival ("survived") with the predictor variables "sex", "age", and "passengerClass".

```{r 1_FitGLM}
# YOUR CODE HERE ####
```



### 2) Model summary 
Output a summary of the model, including coefficients and their significance.

```{r 2_ReviewModel}
# YOUR CODE HERE ####
```



### Follow-up questions:
a) Which predictor variables are statistically significant at the 5% level of significance?
b) How can you interpret the coefficient values? 
c) Which predictor variables are the most materially significant?



# 3) Extra: Interactions
Test some reasonable variable interactions in the model.

```{r 3_Interactions}
# YOUR CODE HERE ####
```

