---
title: 'PPAS Challenge: Practical data concerns'
output: 
  html_document:
    toc: True
  pdf_document:
    toc: True
---

## Background
In this challenge we use a [Kaggle dataset](https://www.kaggle.com/mazharkarimi/heart-disease-and-stroke-prevention/metadata) with data on the prevalence of cardiovascular disease and risk factors. We have created a synthetic, derivative dataset for the purposes of this challenge. The synthetic dataset contains 5 years of seriatim data on heart attack rates by state, year, sex, age, and race.

## Data license
The database license and content license that govern the original dataset can be found in a document in the PPAS GitHub repository. 

## Goals
Prepare data for modeling and validation.

## Load data and packages
```{r Load, warning = F, message = F}
library(dplyr)
library(ggplot2)
library(ROCR)

# Import data and model (from challenge 2) ####
# Only if needed! Feel free to use your own objects from challenge #2.

heartattack <- readRDS("handson_challenges/02_heartdiseasedataset.RDS")
logistic.model <- readRDS("handson_challenges/02_SampleLogisticModel.RDS")
```

## Challenges
### 1) Use the predict() function to create a column of probability predictions in the dataset.

```{r 1_AppendPredictions}

# YOUR CODE HERE ####

```

### 2) Create A/E plots using the holdout subset.

#### a) Across a numeric variable like "Year" (i.e. Year on the x-axis)

```{r 2a_AoverEByYear}

# YOUR CODE HERE ####

```


##### b) Across a categorical variable like "Race" (i.e. Race on the x-axis)

```{r 2b_AoverEByRace}

# YOUR CODE HERE ####

```

### 3) Calculate the AUC on the holdout data.

```{r 3_AUC}

# YOUR CODE HERE ####

```


### 4) Extra: Create a second model and construct a two-way lift chart.

```{r TwoWayLift}

# YOUR CODE HERE ####

```



